---
title: "<span style='font-size: 26px;'>**REGRESSION ANALYSIS**</span><br><span style='font-size: 20px;'>**Quantifying Intervention Methods Impact on Student Math Score Improvement**</span>"
author: "Gareth F"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
  pdf_document:
    toc: true
    latex_engine: xelatex
---

```{css, echo=FALSE}
.author, .date {
  display: none;
}

h2 {
  font-size: 16px;
  font-weight: bold;
  text-transform: uppercase;
}

h3 {
  font-size: 14px;
  font-weight: bold;
}

/* Make main TOC items bold */
.tocify-item {
  font-weight: bold;
}

/* Make ALL sub items consistent */
.tocify-subheader .tocify-item {
  font-weight: normal;
  font-size: 13px;
  padding-left: 20px;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## IMPORT DATA
### Load Packages

```{r load_packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(car)
library(caret)
```

<br>

### Import the 'CSV' file 
```{r import_data}
student <- read.csv("education_intervention_dataset.csv")
```

<br>

### Check for missing or duplicate values in dataset
```{r}
sum(is.na(student))
sum(duplicated(student))
dim(student)
```

<br>

## DATA CLEANING
### Create 'math_improvement' variable
```{r math_improvement}
student$math_improvement <- student$post_math - student$baseline_math
```

<br>

### Label encode 'intervention' variable
```{r label_encode}
student$intervention <- as.numeric(as.factor(student$intervention))
```

<br>

### Remove Irrelevant Columns
```{r remove_columns}
student_math <- student %>%
  select(-baseline_reading, -post_reading, -post_math,
         -student_id, -school_id, -grade_level, -passed)
```

<br>

### Verify Final Dataset
```{r verify_data}
dim(student_math)
colnames(student_math)
head(student_math, 6)
```

<br>

## DATA ANALYSIS 
### Summary of the Data
```{r summary_data}
summary(student_math)
```

<br>

## PLOTTING THE DATA
### Distribution of Math Improvement
```{r eda_histogram, fig.width=10, fig.height=6, fig.align='center'}
ggplot(student_math, aes(x = math_improvement)) +
  geom_histogram(fill = "steelblue", color = "coral", bins = 30) +
  labs(title = "Distribution of Math Score Improvement",
       x = "Math Improvement Score",
       y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

<br>

### Math Improvement by 'Intervention Type'
```{r eda_boxplot, fig.width=10, fig.height=6, fig.align='center'}
ggplot(student, aes(x = factor(intervention), y = math_improvement, 
                    fill = factor(intervention))) +
  geom_boxplot() +
  scale_x_discrete(labels = c("1" = "Blended", "2" = "Control",
                               "3" = "Flipped", "4" = "Tutoring")) +
  labs(title = "Math Score Improvement by Intervention Type",
       x = "Intervention Type",
       y = "Math Improvement",
       fill = "Intervention") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "none")
```

<br>

## REGRESSION MODEL 
### Build the Model
```{r full_model}
model_full <- lm(math_improvement ~ ., data = student_math)
summary(model_full)
```

<br>

### Q-Q Plot of Residuals
```{r qq_plot, fig.width=10, fig.height=6, fig.align='center'}
qqnorm(residuals(model_full),
       main = "Q-Q Plot of Residuals",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles",
       col  = "steelblue",
       pch  = 20)
qqline(residuals(model_full), col = "coral", lwd = 2)
```

<br>

### VIF Multicollinearity Test
```{r vif_test}
vif(model_full)
```

<br>

### Residual Diagnostic Plots
```{r residual_plots, fig.width=10, fig.height=8, fig.align='center'}
par(mfrow = c(2, 2))
plot(model_full, col = "steelblue", pch = 20)
par(mfrow = c(1, 1))
```

<br>

### Stepwise Regression Model
```{r}
model_stepwise <- step(model_full, direction = "both")
summary(model_stepwise)
```


### Cross Validation
```{r cross_validation}
set.seed(123)
train_control <- trainControl(method = "cv", number = 10)
cv_model <- train(math_improvement ~ intervention + 
                  attendance_rate + study_hours_wk + 
                  teacher_experience_years + class_size + 
                  baseline_math + baseline_motivation_1_10,
                  data = student_math,
                  method = "lm",
                  trControl = train_control)
print(cv_model)
cat("Cross Validated R-Squared:", 
    round(cv_model$results$Rsquared, 4), "\n")
cat("Cross Validated R-Squared %:", 
    round(cv_model$results$Rsquared * 100, 2), "%\n")
```

<br>

### Train Test Split
```{r train_test_split}
set.seed(123)
train_index <- createDataPartition(student_math$math_improvement, 
                                   p = 0.80, list = FALSE)
train_data <- student_math[train_index, ]
test_data  <- student_math[-train_index, ]

model_train <- lm(math_improvement ~ intervention + 
                  attendance_rate + study_hours_wk + 
                  teacher_experience_years + class_size + 
                  baseline_math + baseline_motivation_1_10,
                  data = train_data)

predictions <- predict(model_train, newdata = test_data)

ss_res <- sum((test_data$math_improvement - predictions)^2)
ss_tot <- sum((test_data$math_improvement - 
               mean(test_data$math_improvement))^2)
r2_test <- 1 - (ss_res / ss_tot)

cat("Training R-Squared:", round(summary(model_train)$r.squared, 4), "\n")
cat("Test R-Squared:    ", round(r2_test, 4), "\n")
cat("Test R-Squared %:  ", round(r2_test * 100, 2), "%\n")
```

<br>

### Outlier Detection
```{r outlier_check, fig.width=10, fig.height=6, fig.align='center'}
cooks_d <- cooks.distance(model_full)
influential <- which(cooks_d > 4/nrow(student_math))
cat("Number of influential outliers:", length(influential), "\n")
cat("Percentage of dataset:         ", 
    round(length(influential)/nrow(student_math)*100, 2), "%\n")

plot(cooks_d, type = "h",
     main = "Cook's Distance - Outlier Detection",
     xlab = "Observation Index",
     ylab = "Cook's Distance",
     col = "steelblue")
abline(h = 4/nrow(student_math), col = "coral", lwd = 2, lty = 2)
```

<br>

## MODEL EVALUATION
### Extract R-Squared Values
```{r r_squared}
r2_full <- summary(model_full)$r.squared
r2_stepwise <- summary(model_stepwise)$r.squared
```

<br>

### Adjusted R-Squared
```{r adusted_rsquared}
adj_r2_full <- summary(model_full)$adj.r.squared
adj_r2_stepwise <- summary(model_stepwise)$adj.r.squared
```

<br>

### Print Results
```{r print_result}
cat("=== Full Model ===\n")
cat("R-Squared:         ", round(r2_full, 4), "\n")
cat("R-Squared %:       ", round(r2_full * 100, 2), "%\n")
cat("Adjusted R-Squared:", round(adj_r2_full, 4), "\n")

cat("\n=== Stepwise Model ===\n")
cat("R-Squared:         ", round(r2_stepwise, 4), "\n")
cat("R-Squared %:       ", round(r2_stepwise * 100, 2), "%\n")
cat("Adjusted R-Squared:", round(adj_r2_stepwise, 4), "\n")

cat("\n=== Hypothesis Decision ===\n")
if(r2_stepwise > 0.70) {
  cat("R-Squared > 0.70: REJECT H0 - Model exceeds 70% accuracy threshold\n")
} else {
  cat("R-Squared <= 0.70: FAIL TO REJECT H0 - Model does not exceed 70% threshold\n")
}
```







