---
title: "D600: Statistical Data Mining, Linear Regression Analysis"
author: "Gareth Fernes"
date: "`r Sys.Date()`"
output:
  html_document: default
  editor_options:
  chunk_output_type: console
---


# Load Libraries
```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(caret)
library(MASS)
library(leaps)
```


# Import the csv file into R
```{r}
h_info <- read.csv("C:/Users/gfern/OneDrive/Documents/WGU/D600/TASK1/housing_information.csv")
```


# View the file 'housing_info'
```{r}
View(h_info)
```


# Find missing values
```{r}
sum(is.na(h_info))
```


# Inspect column Names
```{r}
colnames(h_info)
```


# Inspect unique values
```{r}
unique(h_info$Windows)
```


# Convert to absolute values
```{r}
h_info$Windows <- abs(h_info$Window)
```


# Calculate summary for 'CrimeRate'
```{r}
mean_crime <- mean(h_info$CrimeRate, na.rm = TRUE)   # Mean
print(paste( "Mean:", round(mean_crime, 2)))
      
median_crime <- median(h_info$CrimeRate, na.rm = TRUE)  # Median
print(paste("Median:", round(median_crime, 2)))

mode_crime <- as.numeric(names(sort(table(h_info$CrimeRate), decreasing = TRUE)[1]))  # Mode
print(paste("Mode:", round(mode_crime, 2)))

range_crime <- range(h_info$CrimeRate, na.rm = TRUE)  # Range
print(paste("Range:", paste(round(range_crime, 2), collapse = " - ")))

min_crime <- min(h_info$CrimeRate, na.rm = TRUE)  # Minimum
print(paste("Minimum:", round(min_crime, 2)))

max_crime <- max(h_info$CrimeRate, na.rm = TRUE)  # Maximum
print(paste("Maximum:", round(max_crime, 2)))

sd_crime <- sd(h_info$CrimeRate, na.rm = TRUE)  # Standard Deviation
print(paste("Standard Deviation:", round(sd_crime, 2)))
```


# Calculate summary for 'DistanceToCityCenter'
```{r}
mean_distance <- mean(h_info$DistanceToCityCenter, na.rm = TRUE)
print(paste("Mean:", round(mean_distance, 2)))

median_distance <- median(h_info$DistanceToCityCenter, na.rm = TRUE)
print(paste("Median:", round(median_distance, 2)))

mode_distance <- as.numeric(names(sort(table(h_info$DistanceToCityCenter), decreasing = TRUE)[1]))  # mode calculation
print(paste("Mode:", round(mode_distance, 2)))

range_distance <- range(h_info$DistanceToCityCenter, na.rm = TRUE)
print(paste("Range:", paste(round(range_distance, 2), collapse = " - ")))

min_distance <- min(h_info$DistanceToCityCenter, na.rm = TRUE)
print(paste("Minimum:", round(min_distance, 2)))

max_distance <- max(h_info$DistanceToCityCenter, na.rm = TRUE)
print(paste("Maximum:", round(max_distance, 2)))

sd_distance <- sd(h_info$DistanceToCityCenter, na.rm = TRUE)
print(paste("Standard Deviation:", round(sd_distance, 2)))
```


# Calculate summary for 'SchoolRating'
```{r}
mean_schoolrating <- mean(h_info$SchoolRating, na.rm = TRUE)
print(paste("Mean:", round(mean_schoolrating, 2)))

median_schoolrating <- median(h_info$SchoolRating, na.rm = TRUE)
print(paste("Median:", round(median_schoolrating, 2)))

mode_schoolrating <- as.numeric(names(sort(table(h_info$SchoolRating), decreasing = TRUE)[1]))  # mode calculation
print(paste("Mode:", round(mode_schoolrating, 2)))

range_schoolrating <- range(h_info$SchoolRating, na.rm = TRUE)
print(paste("Range:", paste(round(range_schoolrating, 2), collapse = " - ")))

min_schoolrating <- min(h_info$SchoolRating, na.rm = TRUE)
print(paste("Minimum:", round(min_schoolrating, 2)))

max_schoolrating <- max(h_info$SchoolRating, na.rm = TRUE)
print(paste("Maximum:", round(max_schoolrating, 2)))

sd_schoolrating <- sd(h_info$SchoolRating, na.rm = TRUE)
print(paste("Standard Deviation:", round(sd_schoolrating, 2)))
```


# Calculate summary for 'TransportAccess'
```{r}
mean_TransportAccess <- mean(h_info$TransportAccess, na.rm = TRUE)
print(paste("Mean:", round(mean_TransportAccess, 2)))

median_TransportAccess <- median(h_info$TransportAccess, na.rm = TRUE)
print(paste("Median:", round(median_TransportAccess, 2)))

mode_TransportAccess <- as.numeric(names(sort(table(h_info$TransportAccess), decreasing = TRUE)[1]))  # mode calculation
print(paste("Mode:", round(mode_TransportAccess, 2)))

range_TransportAccess <- range(h_info$TransportAccess, na.rm = TRUE)
print(paste("Range:", paste(round(range_TransportAccess, 2), collapse = " - ")))

min_TransportAccess <- min(h_info$TransportAccess, na.rm = TRUE)

print(paste("Minimum:", round(min_TransportAccess, 2)))

max_TransportAccess <- max(h_info$TransportAccess, na.rm = TRUE)
print(paste("Maximum:", round(max_TransportAccess, 2)))

sd_TransportAccess <- sd(h_info$TransportAccess, na.rm = TRUE)
print(paste("Standard Deviation:", round(sd_TransportAccess, 2)))
```


# Univariate Visualizations for independent and dependent variables
```{r}
ggplot(h_info, aes(x = CrimeRate)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Crime", x = "Crime Rate", y = "Frequency") +
  theme_minimal()

ggplot(h_info, aes(x = DistanceToCityCenter)) +
  geom_histogram(binwidth = 1, fill = "lightcoral", color = "black") +
  labs(title = "Distribution of Distance To City Center", x = "Distance To City Center", y = "Frequency") +
  theme_minimal()

ggplot(h_info, aes(x = SchoolRating)) +
  geom_histogram(binwidth = 1, fill = "palegreen", color = "black") +
  labs(title = "Distribution of School Rating", x = "School Rating", y = "Frequency") +
  theme_minimal()

ggplot(h_info, aes(x = TransportAccess)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Transport Access", x = "Transport Access", y = "Frequency") +
  theme_minimal()
```


# Bivariate Visualizations for independent and dependent variables
# For 'SchoolRating' and 'CrimeRate'
```{r}
ggplot(h_info, aes(x = CrimeRate, y = SchoolRating, color = SchoolRating)) +
  geom_point(alpha = 0.8, size = 2) +  
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +  # Regression line
  scale_color_gradient(low = "steelblue2", high = "coral") +  
  labs(title = "School Rating by Crime Rate", 
       x = "Crime Rate", 
       y = "School Rating", 
       color = "School Rating") + 
theme_minimal()
```


# For 'SchoolRating' and 'DistanceToCityCenter'  
```{r}
ggplot(h_info, aes(x = DistanceToCityCenter, y = SchoolRating)) +
  geom_point(aes(color = SchoolRating), alpha = 0.8, size = 2) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +  # Add regression line
  scale_color_gradient(low = "palegreen2", high = "lightslateblue") +  
  labs(title = "School Rating by Distance to City Center", 
       x = "Distance to City Center", 
       y = "School Rating", 
       color = "School Rating") + 
  theme_minimal()
```


# For 'SchoolRating' and 'TransportAccess'
```{r}
ggplot(h_info, aes(x = TransportAccess, y = SchoolRating)) +
  geom_point(aes(color = SchoolRating), alpha = 0.8, size = 2) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +  # Add regression line
  scale_color_gradient(low = "lightpink", high = "steelblue") +  
  labs(title = "School Rating by Transport Access", 
       x = "Transport Access", 
       y = "School Rating", 
       color = "School Rating") + 
  theme_minimal()
```


# Next we will select the variables for our analysis
# Define the dataset split and select the variables
```{r}
# Set seed for reproducibility
set.seed(123) 
# Define variables to keep in the dataset
selected_variables <- c("DistanceToCityCenter", "CrimeRate", "SchoolRating", "TransportAccess") 
# Keep only relevant variables
h_info_selected_variables <- h_info[selected_variables]  
```


# Create an index for training data
```{r}
train_index <- createDataPartition(h_info_selected_variables$DistanceToCityCenter, p = 0.7, list = FALSE)
```


# Subset the data into training set and testing set
```{r}
training_data <- h_info_selected_variables[train_index, ]  # 70% Training
test_data <- h_info_selected_variables[-train_index, ]  # 30% Test
```


# Export the test and training datasets to csv files
```{r}
write.csv(training_data, "C:/Users/gfern/OneDrive/Documents/WGU/D600/TASK1/training_data.csv", row.names = FALSE)
write.csv(test_data, "C:/Users/gfern/OneDrive/Documents/WGU/D600/TASK1/test_data.csv", row.names = FALSE)
```


# Check the dimensions of the new dataset
```{r}
dim(training_data)  # Size of training data
dim(test_data)   # Size of testing data
```


# Create a linear regression model with SchoolRating as dependent variable
```{r}
school_rating_model <- lm(SchoolRating ~ DistanceToCityCenter + CrimeRate, data = h_info)
summary(school_rating_model)
```


# Perform stepwise selection
```{r}
stepwise_model <- step(school_rating_model, direction = "both", data = h_info)
summary(stepwise_model)
```


# Find the MSE and run prediction on 'training_data'
```{r}
predictions <- predict(stepwise_model, newdata = training_data)   # Obtain predictions from the optimized model
residuals <- training_data$SchoolRating - predictions   # Calculate residuals
mse <- mean(residuals^2)   # Calculate the Mean Squared Error 
print(paste("Mean Squared Error (MSE):", round(mse, 2)))   # Print the MSE
```


# Run prediction on the 'test_data'
```{r}
predictions <- predict(stepwise_model, newdata = test_data)
actual_values <- test_data$SchoolRating
mse_test <- mean((predictions - actual_values)^2)
print(paste("Mean Squared Error (MSE) on the Test Set:", round(mse_test, 2)))
```


